---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# modendro: more dendro (tree-ring) functions

<!-- badges: start -->
<!-- badges: end -->

modendro is a collection of functions to do a variety of tree-ring analyses that are not available elsewhere or are improvements on existing analyses: various utility functions for wrangling tree-ring data, power transformation & detrending, flexible climate-growth correlations to identify the strongest seasonal climate signals, and disturbance detection & removal.

My inclusion of an analysis method in modendro does not imply broad endorsement of the method. My goal is to make available some of these methods that I've found useful in the past but had to write my own code for. Enjoy and I hope some of these are useful for you too!
  

## Installation

You can install the development version of modendro from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ctmaher/modendro")
```

I recommend you do this often, as modendro is under active development.

## Example: convert rwl-format data.frames to "long" format - and back

The rwl-format data.frame (rownames are years, colnames are series names) is the standard for the important dplR package, and modendro inherits this. However, this format is inconvenient for many other standard operations in R - a "long" format would be much more useful. Enter modendro's `rwl_longer()` function.

```{r example 1.1}
library(modendro)
data("PerkinsSwetnam96")
head(PerkinsSwetnam96)[,1:5] # typical rwl-format


PS.long <- rwl_longer(rwl = PerkinsSwetnam96,
                      series.name = "series", # name the series IDs whatever you want
                      dat.name = "rw.mm", # same for the measurement data
                      trim = TRUE, # trim off the NAs before and after a series
                      new.val.internal.na = NULL) # leave NAs internal to the measurement series
# Note that we could replace internal NAs here if there were any (e.g., with 0s)

head(PS.long) # familiar format

```

Once our tree-ring data is in the familiar (to me anyway) long format, we do all kinds of data wrangling, analyses and plotting that we normally do in R. For example, we might want to merge the tree-ring data with site IDs so that we can add site as a random effect in a model or for plotting.

```{r example 1.2}

data("PSgroupIDs")

PS.long.sites <- merge(PS.long, PSgroupIDs, by = "series")

library(ggplot2)

ggplot(PS.long.sites, aes(year, rw.mm, group = series)) +
  geom_line(alpha = 0.5) +
  facet_wrap( ~ site, ncol = 1)

```

Another useful application of `rwl_longer()` is when we have several rwl-format files that we want to bind together. This is not intuitive with the rwl-format - the number of rows may not line up and the columns won't either. This is simple with the long format. 

We'll borrow some data from the dplR package for this.

We may want to convert this joined data back into a single rwl-format data.frame, e.g., for export or for unifying analyses in modendro or dplR. We can do this with modendro's `longer_rwl()`.

```{r example 1.3}

library(dplR)
data("ca533")

ca533.long <- rwl_longer(ca533,
                         series.name = "series",
                         dat.name = "rw.mm")

comb.long <- rbind(PS.long, ca533.long)

# note that longer_rwl() doesn't care if you have other variables in your long data.frame
# We'll add some here to illustrate that
comb.long$foo <- "bar"
comb.long$bar <- "foo"

comb.rwl <- longer_rwl(df = comb.long,
                       series.name = "series",
                       dat.name = "rw.mm")

head(comb.rwl)[,1:5]

```


## Example: power transformation & detrending ala Cook & Peters (1997)



