---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# modendro: more dendro (tree-ring) functions

<!-- badges: start -->
<!-- badges: end -->

  modendro is a collection of functions to do a variety of tree-ring analyses that are not available elsewhere or are improvements on existing analyses: power transformation 
& detrending, disturbance detection, flexible climate-growth correlations to identify the strongest seasonal climate signals, & various utility functions for wrangling tree-ring data. 
  The functions in modendro are designed to work with the common tree-ring measurement formats that have years as rownames and series names as colnames (i.e., the resulting format from the `read.` functions in the dplR package). Within R, this format is just a particularly-structured data.frame. The modendro functions don't "know" what file format the measurements originated from, and the only things that matter are the rownames and the colnames

## Installation

You can install the development version of modendro from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ctmaher/modendro")
```

## Example: convert rwl-format data.frames to "long" format - and back

The rwl-format data.frame is the standard for the important dplR package, and modendro inherits this. However, this format is inconvenient for many other standard operations in R - a "long" format would be much more useful. Enter modendro's `rwl_longer()` function

```{r example 1}
library(modendro)
data("PerkinsSwetnam96")
head(PerkinsSwetnam96) # typical rwl-format


# PS.long <- rwl_longer(rwl = PerkinsSwetnam96,
#                       series.name = "series", # name the series IDs whatever you want
#                       dat.name = "rw.mm", # same for the measurement data
#                       trim = TRUE, # trim off the NAs before and after a series
#                       new.val.internal.na = NULL) # leave NAs internal to the measurement series
# # Note that we could replace internal NAs here if there were any (e.g., with 0s)
# 
# head(PS.long) # familiar format

```

Once our tree-ring data is in the familiar (to me anyway) long format, we do all kinds of data wrangling, analyses and plotting that we normally do in R. For example, we might want to merge the tree-ring data with site IDs so that we can add site as a random effect in a model or for plotting.

```{r example 2}

data("PSgroupIDs")

# PS.long.sites <- merge(PS.long, PSgroupIDs, by = "series")
# 
# library(ggplot2)
# 
# ggplot(PS.long.sites, aes(year, rw.mm)) +
#   geom_line(alpha = 0.5) +
#   facet_wrap( ~ site)

```


More to come...
